Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-12-29 concurrency/Verification of Concurrent Binary Search Tree.tex
updated todo list
started describing lock invariants and global ghost state

\newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
%\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
\newcommand{\islock}{\boxdotright}
\newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
\newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\emp}{\mathsf{emp}}

@@

We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations. The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section. From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock. These pieces of ghost state can then be composed to yield the abstract state of the entire data structure. In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \islock R(a)$, where the value of $a$ can be updated in the critical section as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole. This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.

The construction of the lock variants is simple. We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$). When the share $\pi$ is 1, we know that $p = a$. (do we want to define this ourselves, or reference it somewhere? note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
$$\ell \islock R(a) \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure. Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change. Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure. %smaller example before we get into BST?

@@

To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. Our overall approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

\subsubsection{Ghost States and Range}
 The main challenge in proving the properties of concurrent data structure is to abstract the data structure in a way that helps to reason about it in the concurrent setting. Krishna et al. \cite{krishna2017flow} mention in their paper that we can abstract binary search tree using appropriate bounds in each node: we can use the product of \emph{lower-bound} and \emph{upper-bound} on each node of binary tree by propagating from each node to its child the appropriate bounds on the values presents in the sub-tree. The \emph{lower-bound} and \emph{upper-bound} help us to prove that we perform the operation in the right node: for instance, if we $\mathsf{insert}$ a key in the tree, then key should be inside those two bounds. We use the term \emph{range} to represent the product of lower- and upper-bound. \ref{range_bst} shows an example of BST with each node labeled with appropriate bound propagated all the way from \emph{root} to empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
The range for root node is always $(-\infty,+\infty)$ and then, we can calculate the range for all nodes in the tree by enforcing each node to propagates appropriate bounds. For the left child, range will be lower-bound for the node (lower-bound) and the key in the node (upper-bound), while for the right child range will be a key in the node (lower-bound) and upper-bound for the node (upper-bound). This properties in each node help us to prove that the thread have changed the state of data-structure to the valid state. For instance, we can confirm that a key has been inserted in the right place if a key is inside the range of the node. Suppose we want to insert a key $38$ in \ref{range_bst}, then we will locate \emph{right} child of $35$ as the right place for $38$. During insertion, we can confirm that the node is indeed right place for $38$ by checking that $38$ is in the range (35,40). We introduce the $range$ in our proof using the ghost variables. This $ranges$ along with other information about the node represent the abstract state of the data structure. We have used two kind of ghost state in our proof: \emph{per-node} ghost state and \emph{overall} ghost state.\\
\textbf{Per-node Ghost State}: The per-node ghost state is includes a node's range and other information about the node: its current key, value, and ghost names of its child nodes. In VST, we can use a type as ghost state by proving an instance of the \texttt{Ghost} typeclass, analogous to the \emph{resource algebras} of Iris. The essential part is a \emph{join} relation describing how two pieces of ghost state combine. We define the join of two ranges to be the range that contains both of them, and use an \emph{exclusive} algebra for the rest of the data, so that only one thread knows the current state of a node at a time. 
Ghost state is shared between the per-node lock invariants and the abstract state of the data structure using the \emph{reference} pattern: each thread holds partial information describing its contribution to the shared state, and the shared resources hold a ``reference'' copy that records all of the contributions. In VST, this is implemented with assertions $\mathsf{my\_half}\ g\ \mathit{sh}\ a$ and $\mathsf{public\_half}\ g\ r$, where $g$ is the node's identifier ($\gnamety$), $a$/$r$ is the value of the ghost state, and $\mathit{sh}$ is a share describing how much of the reference value is known: if $\mathit{sh}$ is the full share, then we know that $a = r$. In the proof, the \texttt{my\_half} assertion for a node will be held by the node's lock, and \texttt{public\_half} will be held by the shared state.

\textbf{Overall Ghost State}: We now have per-node ghost state for each node, indexed by a ghost name $g_n$. But if we are given an arbitrary $\gnamety$ for a node, we do not know whether that $\gnamety$ exists in the tree. We need another kind of ghost state to hold the set of all nodes in the current tree. Again we can use the ``reference'' pattern to define this ghost state, where the abstract state $\mathsf{ghost\_ref}\ s$ holds the full set of nodes, and each node's lock invariant holds an $\mathsf{in\_tree\ g_n}$ assertion guaranteeing that $g_n$ is in $s$.%notation for gnames
\ignore{\begin{verbatim}
Definition ghost_ref g r1 := ghost_reference(P := set_PCM) r1 g.
Definition in_tree g g1 := EX sh: share, ghost_part(P :=set_PCM) sh (Ensembles.Singleton g1) g.
\end{verbatim}}
\ignore{The global invariant, which we introduced in our atomic specification as the $\treerep$ predicate above, ties together the pieces as follows:
\begin{align*}\treerep(T,g) \triangleq *_n. \mathsf{public\_half\textsubscript{gn}} \ (\mathit{range}_n* \mathit{ghost\_info}_n)\ ] *\ \mathsf{ghost\_ref}\ g\ \{ gn\  \} \end{align*}
 Later while writing atomic specification, we will use $\mathsf{ghost\_ref}$ and $\mathsf{in\_tree}$ in the public and private part respectively. Once we open \emph{global invariant}, we get access to $\mathsf{ghost\_ref}$ which will be used with $\mathsf{in\_tree}$ to establish the fact that node actually exists in the tree. We proved the following lemma for this:
$$
Lemma\ \mathbf{\ node\_exist\_in\_tree : }\ \forall\ g\ s\ g\_in,\  \mathbf{in\_tree}\ g\ g\_in  * \mathbf{ghost\_ref}\ g\ s\ \Rightarrow\ g\_in\ \in s .
$$}
\subsubsection{Handle and Abstract State}
Now we are ready to define the $\treerep$ and $\nodeboxrep$ predicates.

describe lock invariant

We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

@@

>>>>

\newcommand{\TODO}[1]{\textbf{\textcolor{red}{[ TODO: #1]}}}
%\newcommand{\boxdotright}{\!\mathrel\boxdot\joinrel\rightarrow\!}
\newcommand{\islock}{\boxdotright}
\newcommand{\lockvar}{\islock^{v}}
\newcommand{\isaex}{\!\mathrel\odot\joinrel\rightarrow\!}
\newcommand{\xisaex}[1]{\!\mathrel\odot\joinrel\xrightarrow{#1}\!}
%% \newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\emp}{\mathsf{emp}}

@@

We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations. The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section. From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock. These pieces of ghost state can then be composed to yield the abstract state of the entire data structure. In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \lockvar R$, where $R$ is parameterized by a value that can be updated in the critical section, as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole. This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.

The construction of the lock variants is simple. We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$). When the share $\pi$ is 1, we know that $p = a$. (do we want to define this ourselves, or reference it somewhere? note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure. Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change. Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure. %smaller example before we get into BST?

@@

To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. As described above, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

\subsubsection{Key Ranges}
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by \citet{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
At the leaves, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

\subsubsection{Per-Node and Global Ghost State}
Importantly, the fact that the leaf ranges partition the space of possible keys remains true even if operations in other threads restructure the tree during our traversal. Furthermore, operations further up the tree can only \emph{increase} the range of nodes below them, so if we find a node whose range includes our key, we are guaranteed to have found the right place for it. This makes ranges ideal for use as ghost state: range information held by one thread will not be invalidated by other threads, and is sufficient to show correctness of BST operations. More precisely, the abstract state of a node in the tree consists of its range plus a representation of its contents: for a leaf node, $\mathsf{None}$, and for an internal node, $\mathsf{Some}$ of its current key, value, and the identifiers of its left and right children. This per-node abstract state serves as a bridge between the per-node locks of the implementation and the abstract state of the entire tree (the $\treerep$ predicate) in the specification.

The abstract state of a tree is a map from keys to values. The $\treerep$ predicate then describes the conditions under which a map $m$ is implemented by the ghost state of a tree. Roughly speaking, we should be able to gather the per-node ghost states into an abstract tree $t$ that implements the map $m$, in that each key-value pair in $m$ appears in one of the nodes of $t$. We formalize this by first defining the collection of $\mathsf{public\_half}$ assertions---i.e., the authoritative copies of the per-node ghost states---needed to form an abstract tree:
\begin{align*}
\mathsf{public}&\mathsf{\_halves}(t, g, (l, r)) \triangleq \mathsf{match}\ t\ \mathsf{with}\\
& |\ \mathsf{Leaf} \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{None})\\
& |\ \mathsf{Node}\ (k, v, t_l, g_l, t_r, g_r) \Rightarrow \mathsf{public\_half}_g\ ((l, r), \mathsf{Some}\ (k, v, g_l, g_r)) * \mathsf{public\_halves}(t_l, g_l, (l, k)) * \mathsf{public\_halves}(t_r, g_r, (k, r))
\end{align*}
We recursively walk through the tree, starting from a node $g$ with range $(l, r)$, and collect the $\mathsf{public\_half}$ assertions of each node into a single assertion. We compute the range of each node's children as described above: if the key in the current node is $k$, then the left child has range $(l, k)$ and the right child has range $(k, r)$.

This collection of $\mathsf{public\_half}$ assertions can be combined with per-node lock variants to ensure that the nodes in an implementation actually form a valid tree. However, we also need to record the fact that each node in the implementation actually belongs to the tree $t$. We do this by augmenting $\mathsf{public\_halves}$ with a piece of ghost state $\mathsf{ghost\_nodes}(S)$ that records the set of all nodes in the tree, constructed so that each node can hold an $\mathsf{in\_tree}\ g$ assertion guaranteeing that $g \in S$. Then our final definition of $\treerep$ is:
$$\treerep(m, g) \triangleq \exists t.\ t \text{ implements } m \land \mathsf{public\_halves}(t, g, (-\infty, +\infty)) * \mathsf{ghost\_nodes}(\mathsf{node\_names}(t))$$
The range of the root node (named $g$) is $(-\infty, +\infty)$, and the $\mathsf{ghost\_nodes}$ assertion guarantees that if any thread or lock holds an assertion $\mathsf{in\_tree}\ g_i$, then $g_i$ is a node in the ghost tree $t$. Because we quantify over the abstract tree $t$, we can rearrange its nodes at any time, as long as it remains a well-formed tree with the same set of node names and key-value pairs. (We will take advantage of this property when verifying the \lstinline{delete} function.)

As described in section~\ref{atomicity}, we connect the per-node locks of our C implementation with the global abstract state of $\treerep$ by associating each node's lock with a lock variant assertion $\ell \islock_\pi R$, where $R$ is parameterized by the abstract state of the node (range + possible contents). We will fix the share $\pi$ to 1/2 so that the range in a node's lock can be out of date with respect to its $\mathsf{public\_half}$ (in particular, it may have grown due to rearrangements further up the tree). We then need to choose a variant $R$ describing the resources held by the lock for a given abstract state. The C implementation of a node is:
\begin{lstlisting}[language = C,numbers = none]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
\end{lstlisting}
The node data of each \lstinline{tree} is wrapped in a \lstinline{tree_t} struct, and is only accessible when the lock in the \lstinline{lock} field is held. The lock variant for a node at location $p$ then needs to describe the contents of the \lstinline{t} field for each possible abstract state:
\begin{align*}
&\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r)
\end{align*}
Because our implementation is doing hand-over-hand locking, in the internal case, we also need to obtain the lock variant assertions for the child nodes, so the full lock variant is: %this isn't enough; we need to talk about selflock and shares
\begin{align*}
\mathsf{node\_inv}((l, r), c) \triangleq \ &\mathsf{match}\ c\ \mathsf{with}\\
&|\ \mathsf{None} \Rightarrow \texttt{t} \mapsto \texttt{NULL}\\
&|\ \mathsf{Some}\ (k, v, g_l, g_r) \Rightarrow k \in (l, r) \land \exists p_l, p_r.\ \texttt{t} \mapsto (k, v, p_l, p_r) * p_l \lockvar \mathsf{node\_inv} * p_r \lockvar \mathsf{node\_inv}
\end{align*}
Because we are using lock variants, we do not need to know the ranges or contents of the child nodes: instead, the variant construction guarantees that they are consistent with the global ghost state.

%%edit from here
We combine pieces of information discussed above, \emph{recursive lock invariant} from \ref{safety} and \emph{ghost states}, to write the atomic specification. The private pre- and postcondition of the atomic specification is given by defining the predicate $\nodeboxrep(p,g,g\_root)$ that describes the concrete state of the BST along with the contribution part of the both kind of ghost states explained above.The root node $p$ is represented by ghost name $g\_root$ and overall ghost state is represented by ghost name $g$. The $\nodeboxrep$ predicate ties together $\mathsf{my\_half}$ part of per-node ghost state and $\mathsf{in\_tree}$ of over-all ghost state inside the $\mathsf{lock\_inv}$ assertion of each node's lock as follows:
\begin{align*}
    &\nodeboxrep_{\pi}(p,g,g\_root) \triangleq  p\mapsto\ (lock,tp)\ *\ \mathsf{lock\_inv_{\pi}}(lock, \mathsf{selflock}_{0.5}(lock,R))) \\&R  \triangleq\ \ghost{\mathsf{my\_half}(range,info)}{g\_root}\ *\ \ghost{\mathsf{in\_tree}\ (g\_root)}{g}\ *\ \exists\ pa\ pb\ ga\ gb,\ tp\mapsto\ (pa * pb)\ *\  \nodeboxrep_{0.5}(pa,g,ga) \\& *\ \nodeboxrep_{0.5}(pb,g,gb))\end{align*}

@@

\section{Using the Specifications}
client and client proofs
